### 2024-08-29 player-evaluation.qmd
20240828 - additions
20240829 - signs and samples

Compared to tests of speed, power, and agility at the NFL Combine, collegiate performance is a stronger predictor of performance in the NFL [@Lyons2011].
That is, previous sports performance is the best predictor of future performance [for a review, see @DenHartigh2018].
Thus, it is important to consider a player's past performance.
However, the extent to which historical performance may predict future performance may depend on many factors such as (a) the similarity of the prior situation to the current situation, (b) how long ago the prior situation was, and (c) the extent to which the player (or situation) has changed in the interim.
For rookies, the player does not have prior seasons of performance in the NFL to draw upon.
Thus, when evaluating rookies, it can be helpful to consider their performance in college or in their prior leagues.
However, there are large differences between the situation in college and the situation in the NFL, so prior success in college may not portend future success in the NFL.
An indicator that intends to be prognostic of future performance, and that accounts for past performance, is a player's draft position—that is, how early (or late) was a player selected in the NFL Draft.
The earlier a player was selected in the NFL Draft, the greater likelihood that the player will perform well; however, this is somewhat countered by the fact that the teams with the highest draft picks tend to be the worst based on the prior season's record.

@@

In addition to how many fantasy points a player scores in terms of historical performance, we also care about efficiency and [consistency](#sec-evalHistoricalConsistency).
How efficient were they given the number of opportunities they had?
If they were relatively more efficient, they will likely score more points than many of their peers when given more opportunities.
If they were relativelly inefficient, their capacity to score fantasy points may be more dependent on touches/opportunities.
Efficiency might be operationalized by indicators such as yards per passing attempt, yards per rushing attempt, yards per target, yards per reception, etc.

@@

@@

After performing an evaluation of the relevant domain(s) for a given player, then one must integrate the evaluation information across domains to make a judgment about a player's overall value.
When thinking about a player's value, it can be worth thinking of a player's upside and a player's downside.
Player that are more consistent may show higher downside but a lower upside.
Younger, less experienced players may show a higher upside but a lower downside.

>>>>

It is important to consider a player's past performance.
However, the extent to which historical performance may predict future performance may depend on many factors such as (a) the similarity of the prior situation to the current situation, (b) how long ago the prior situation was, and (c) the extent to which the player (or situation) has changed in the interim.
For rookies, the player does not have prior seasons of performance in the NFL to draw upon.
Thus, when evaluating rookies, it can be helpful to consider their performance in college or in their prior leagues.
However, there are large differences between the situation in college and the situation in the NFL, so prior success in college may not portend future success in the NFL.
An indicator that intends to be prognostic of future performance, and that accounts for past performance, is a player's draft position—that is, how early (or late) was a player selected in the NFL Draft.
The earlier a player was selected in the NFL Draft, the greater likelihood that the player will perform well; however, this is somewhat countered by the fact that the teams with the highest draft picks tend to be the worst based on the prior season's record.

@@

In addition to how many fantasy points a player scores in terms of historical performance, we also care about efficiency and [consistency](#sec-evalHistoricalConsistency).
How efficient were they given the number of opportunities they had?
This is important to consider because different players have different opportunities.
For example, some Running Backs had more opportunity than others (e.g., more carries), so comparing players merely on rushing yards would be misleading.
If they were relatively more efficient, they will likely score more points than many of their peers when given more opportunities.
If they were relativelly inefficient, their capacity to score fantasy points may be more dependent on touches/opportunities.
Efficiency might be operationalized by indicators such as yards per passing attempt, yards per rushing attempt, yards per target, yards per reception, etc.

@@

## Signs Versus Samples {#sec-signsSamples}

When thinking about the most important domains to assess for evaluating players and predicting their performance, it is important to distinguish between signs and samples [@DenHartigh2018].
*Signs* are indicators of underlying states.
The signs approach to assessment involves assessing processes that may predict performance, where the emphasis is on what the sign indicates about the underlying attribute, rather than on the specific behavior itself.
In football, the signs approach might involve measuring skills using separate tests where players' skills are tested in isolation.
For instance, the NFL Combine takes a signs approach to assessment, in which players have a separate test to assess for speed (e.g., 40-yard dash) and jumping ability (e.g., vertical jump).

*Samples* reflect behaviors that are close to the behavior of interest.
The samples approach to asesssment tries to assess the person's performance in a behavior, situation, and context that are as reflective as possible of the kinds of behaviors, situations, and contexts that the person would have to perform in for the position or occupation.
In the samples approach, the behavior itself is the main emphasis, rather than an underlying attribute.
In football, the samples approach might involve measuring the player's performance during games or game-like situations.
For instance, to assess the skills of a Wide Receiver, you might observe—during games or game-like situations—how well they are able to catch passes when closely guarded or double-teamed, or how well they are able to catch poorly thrown passes.

In general, samples tend to be stronger than signs for predicting player performance in sports [@DenHartigh2018].
For instance, compared to tests of speed, power, and agility at the NFL Combine, collegiate performance is a stronger predictor of performance in the NFL [@Lyons2011].
That is, previous sports performance is the best predictor of future performance [for a review, see @DenHartigh2018].

According to the theoretical perspective to sports performance known as the ecological dynamics approach, successful performance in sports involves the coordination of multiple, intertwined skills that are contextually embedded [@DenHartigh2018].
For instance, a Wide Receiver needs to coordinate skills in speed, route running, jumping, and good hands to create space from defenders and to catch poorly thrown passes in a game.
The ecological dynamics approach considers the important interaction of the player, task, and environment.
Thus, to assess players in a way that is likely to be most predictive of their future performance, it is important for the assessment to retain the player–task–environment interaction [@DenHartigh2018].
For example, it would be valuable to use assessments from games or game-like situations in which players must perform tasks that leverage multiple, intertwined skills and that are similar to the tasks you want to predict.

@@

After performing an evaluation of the relevant domain(s) for a given player, one must integrate the evaluation information across domains to make a judgment about a player's overall value.
When considering how much weight to give to each of various factors, it is important to evaluate how much [predictive validity](#sec-predictiveValidity) each factor has for predicting a player's successful performance in the NFL.
Then, one can weight each variable according to its [predictive validity](#sec-predictiveValidity) using an [actuarial approach](#sec-actuarialPrediction).
[Actuarial approaches](#sec-actuarialPrediction) are described in a later chapter, in @sec-actuarialPrediction.
For now, suffice it to say that [actuarial approaches](#sec-actuarialPrediction) leverage statistical formulas, as opposed to using [judgment](#sec-humanJudgment) alone.
As described in @sec-cognitiveBias, people's judgment—including judgments by experts (e.g., professional scouts and coaches)—tends to be riddled with biases [@DenHartigh2018].
Moreover, experts tend to disagree in terms of their judgments/predictions about players [@DenHartigh2018].
In addition, there are benefits to leveraging multiple perpectives, consistent with the "[wisdom of the crowd](#sec-wisdomOfCrowd)", as described in @sec-wisdomOfCrowd.
In general, as noted in @sec-signsSamples, give more weight to [samples](#sec-signsSamples) of relevant behavior—such as past performance in games—than to [signs](#sec-signsSamples) such as NFL Combine metrics (e.g., bench press, 40-yard dash, etc.).
Where you can, use assessments from games or game-like situations in which players must perform tasks that leverage multiple, intertwined skills and that are similar to the tasks you want to predict.

When thinking about a player's value, it can be worth thinking of a player's upside and a player's downside.
Players that are more consistent may show higher downside but a lower upside.
Younger, less experienced players may show a higher upside but a lower downside.
